# Class Activity 7


## Your Turn 1

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, comment=NA, warning = FALSE, message = FALSE)
```

## Few Instructions

The default setting in Rstudio when you are running chunks is that the “output” (numbers, graphs) are
shown **inline** within the Markdown Rmd. If you prefer to have your plots appear on the right of the console and not below the chunk, then change the settings as follows:

1. Select Tools > Global Options.
2. Click the R Markdown section and uncheck (if needed) the option Show output inline for all
R Markdown documents.
3. Click OK.

Now try running R chunks in the .Rmd file to see the difference. You can recheck this box if you prefer
the default setting.

## Parameters and Statistics 
Here are some notations that will be useful for you. Look for the codes to produce this in the associated Rmd file.

&nbsp; | \textcolor{red}{Parameter} | \textcolor{red}{Statistic} |
---- | ---- | ---- |
Mean | $\mu$ | $\bar{x}$ | 
Proportion | $p$ | $\hat{p}$ |
Std. Dev. | $\sigma$ | $s$ |
Correlation | $\rho$ | $r$ |
Slope | $\beta$ | $b$ |

\vspace*{1in}

## Example 1: Parameters and Statistics

For each of the following, state whether the quantity described is a parameter or a statistic, and give the correct notation.

### a. Average household income for all houses in the US, using data from the US census

*Answer:* This is a parameter since the mean is for all houses in the US, and the notation is $\mu$.

### b. The proportion of all residents in a county who voted in the last presidential election.

*Answer:* This is a parameter since we have information on all the residents, and the notation is p.

### c. The difference in proportion who have ever smoked cigarettes, between a sample of 500 people who are 60 years old and a sample of 200 people who are 25 years old.    

*Answer:* We use statistics since the proportions are from samples.  The notation for the difference in sample proportions is $\hat{p}_1 - \hat{p}_2$

### d. The correlation between weight and height for 5-year old kids. 

*Answer:* If we are looking at all 5-year old kids it is a parameter, and the notation for correlation is $\rho$.

### e. The mean number of extracurricular activities from a random sample of 50 students at your school.

*Answer:* This is a statistic since the mean is from a sample, and the notation is $\mu$.

\vspace*{1in}

## Example 2: Using Search Engines on the Internet

A 2012 survey of a random sample of 2253 US adults found that 1,329 of them reported using a search engine (such as Google) every day to find information on the Internet.  

### a).  Find the relevant proportion and give the correct notation with it.  

*Answer:* $\hat{p} = 1329/2253$

```{r}
p.hat <- 1329/2253
p.hat
```


### b).  Is your answer to part (a) a parameter or a statistic?  

*Answer:* Statistic

### c).  Give notation for and define the population parameter that we estimate using the result of part (a).

*Answer:* p = the proportion of all US adults that would report that they use an Internet search engine every day

\vspace*{1in}

## Example 3: Simulation of a Sample Proportion

According to a PEW survey, $66\%$ of U.S. adult citizens casted a ballot in the 2020 election. Suppose we take a random sample of $n=100$ eligible U.S. voters and computed the sample proportion who voted. 


```{r}
# Call the library
library(ggplot2)
```


```{r}
# Define parameters
pop.prop <- .66 # Population proportion
n.size <- 100  # sample size
```


### (a) Generate a random sample of size $n= 100$ and plot its sample proportion.

```{r}
# Generate 1 sample
sample1 <- rbinom(n = 1, size = n.size, p = pop.prop) # R simulates the samples
sample.prop1 <- sample1/n.size #  Proportion = No. of Success / Sample Size

```


```{r}
# define a data frame
mydata <- data.frame(x = sample.prop1)

# Plot a dot plot of the sample proportion
ggplot(mydata, aes(x = sample.prop1)) +
  geom_dotplot(dotsize=0.25, stackratio=0.75, binwidth=0.01) +
  ggtitle("A single sample proportion") +  xlab("Proportion") + ylab("Count")+
  scale_x_continuous(limits = c(0.4, 0.9))+
  theme(plot.title = element_text(hjust = 0.5))

```

### (b) Generate 5 random samples of size $n= 100$ and plot the sample proportions.


```{r}
# generate 5 random samples of size 100
sample5 <- rbinom(n = 5, size = n.size, p = pop.prop)  
sample.prop5 <- sample5/n.size 

data <- data.frame(x = sample.prop5)

ggplot(data, aes(x = sample.prop5)) +
  geom_dotplot(dotsize=0.25, stackratio=0.9, binwidth=0.01 )+
  ggtitle("") +  xlab("Proportion") + ylab("Count")+
  scale_x_continuous(limits = c(0.4, 0.9))+ 
  theme(plot.title = element_text(hjust = 0.5))

```


### (c) Generate 1000 random samples of size $n= 100$ and plot the sample proportions.

```{r}
# Generate 1000 samples
sample1000 <- rbinom(n = 1000, size = n.size, p = pop.prop)  
sample.prop1000 <- sample1000/n.size 

data <- data.frame(x = sample.prop1000)

ggplot(data, aes(x = sample.prop1000)) +
  geom_dotplot(dotsize=0.25, method = "histodot", stackratio=0.9, binwidth=0.01) +
  ggtitle("") +  xlab("Proportion") + ylab("Count") +
  scale_x_continuous(limits = c(0.4, 0.9))+ 
  theme(plot.title = element_text(hjust = 0.5))

```

*Question:* What does each dot represent?

*Answer:* One sample proportion from a sample of n=100 eligible voters.

 *Question:* What is the shape of your sampling distribution?

*Answer:* Roughly symmetric.

*Question:* Where is your distribution centered?

*Answer:* About 0.66, which is the population proportion. 

*Question:* The distribution should be centered at the population proportion. Verify that the distribution is centered around the population proportion, $p = 0.66$. 

*Answer:* 

```{r}
# r-code
mean(sample.prop1000)
```


*Question:* What is the standard deviation of this distribution? (Hint: use the 95% rule.) 

*Answer:* About 0.03, it looks like most sample proportions are between 0.55 to 0.75 so 2 standard deviations is about 0.10. This makes the SD about 0.05. 

*Question:* The standard deviation of sampling distribution has a separate name. It is called the **Standard Error**. Verify the standard deviation of this distribution using R-code.

*Answer:*

```{r}
# r-code
sd(sample.prop1000)
```


### (d) Repeat part(c) with sample size 20 instead of 100. Generate 1000 samples.

```{r}
# Generate 1000 samples
n.size <- 20

sample1000 <- rbinom(n = 1000, size = n.size, p = pop.prop)  
sample.prop1000 <- sample1000/n.size 

data <- data.frame(x = sample.prop1000)

ggplot(data, aes(x = sample.prop1000)) +
  geom_dotplot(dotsize=0.225, method = "histodot", stackratio=0.8, binwidth=0.01) +
  ggtitle("") +  xlab("Proportion") + ylab("Count") +
  scale_x_continuous(limits = c(0.3, 0.9))+ 
  theme(plot.title = element_text(hjust = 0.5))

```


*Question:*  How has the sampling distribution changed? (Shape? Center? Variability?)

*Answer:* The shape is slightly left skewed, still centered at 0.66 but with more variability that before (SD of about 0.10). This distribution is more discrete looking because there are just a few sample proportions possible with n=20 (e.g. 20/20, 19/20, 18/20, etc). 

```{r}
mean(sample.prop1000)
sd(sample.prop1000)
```


### (d) Now suppose the population proportion is $p=0.90$ instead of $p=0.66$ in part (e). Keep n.size=20. 

```{r}
# Generate 1000 samples

pop.prop <- 0.90
n.size <- 20
n.size <- 20

sample1000 <- rbinom(n = 1000, size = n.size, p = pop.prop)  
sample.prop1000 <- sample1000/n.size 

data <- data.frame(x = sample.prop1000)

ggplot(data, aes(x = sample.prop1000)) +
  geom_dotplot(dotsize=0.21, method = "histodot", stackratio=0.8, binwidth=0.01) +
  ggtitle("") +  xlab("Proportion") + ylab("Count") +
  scale_x_continuous(limits = c(0.4, 0.9))+ 
  theme(plot.title = element_text(hjust = 0.5))

```

*Question:* How has the sampling distribution changed? (Shape? Center? Variability?)

*Answer:*  The shape is much more left skewed than when p=0.66. Center is around 0.90 and SD is around 0.07. Note that increasing the population proportion closer to 1 results in a decrease in the SD because most samples give proportion near 1.  

```{r}
mean(sample.prop1000)
sd(sample.prop1000)
```


## Example 4: Simulation for a Sample Mean

We’ll look at sampling movies from the population of 134 Hollywood movies made in 2011 and measuring their budget (millions of dollars).

```{r}
# import dataset
library(Lock5Data)
movies <- HollywoodMovies2011
```

### (a) What is the population mean of the `Budget`?

```{r}
# r-code
mean(movies$Budget, na.rm = TRUE)
```

### (b) Generate a random sample of size $n= 10$ and plot the sample proportion.

```{r}
# define a data frame
n.size <- 10

Budget <- movies$Budget[!is.na(movies$Budget)] # remove NAs

sample1 <- sample(Budget, size = n.size) 
sample.mean1 <- mean(sample1)

mydata <- data.frame(x = sample.mean1)

# Plot a dot plot of the sample proportion
ggplot(mydata, aes(x = sample.mean1)) +
  geom_dotplot(dotsize=1, stackratio=0.9, binwidth=1) +
  ggtitle("A single sample mean") +  xlab("Sample mean") + ylab("Count")+
  scale_x_continuous(limits = c(1,120))+ 
  theme(plot.title = element_text(hjust = 0.5))


```


### (c) Generate 5 random samples of size $n= 10$ and plot the sample means.

```{r}
n.size <- 10
n.rep <- 5

Budget <- movies$Budget[!is.na(movies$Budget)] # remove NAs

sample5 <- lapply(1:5, function(i) sample(Budget, size = n.size)) 
sample.mean5 <- lapply(sample5, function(x) mean(x)) 
sample.mean5 <- unlist(sample.mean5)

mydata <- data.frame(x = sample.mean5)

# Plot a dot plot of the sample proportion
ggplot(mydata, aes(x = sample.mean5)) +
  geom_dotplot(dotsize=0.9, stackratio=0.9, binwidth=1) +
  ggtitle("A single sample mean") +  xlab("Sample mean") + ylab("Count")+
  scale_x_continuous(limits = c(1,120))+ 
  theme(plot.title = element_text(hjust = 0.5))

```


### (d) Generate 1000 random samples of size $n= 10$ and plot the sample means.

```{r}
# Generate 1000 samples
n.size <- 10
n.rep <- 1000

Budget <- movies$Budget[!is.na(movies$Budget)] # remove NAs

sample1000 <- lapply(1:n.rep, function(i) sample(Budget, size = n.size)) 
sample.mean1000 <- lapply(sample1000, function(x) mean(x)) 
sample.mean1000 <- unlist(sample.mean1000)

mydata <- data.frame(x = sample.mean1000)

# Plot a dot plot of the sample proportion
ggplot(mydata, aes(x = sample.mean1000)) +
  geom_dotplot(dotsize=1, method = "histodot", stackratio=0.9, binwidth=1) +
  ggtitle("A single sample mean") +  xlab("Sample mean") + ylab("Count")+
  scale_x_continuous(limits = c(1,120))+ 
  theme(plot.title = element_text(hjust = 0.5))
```


*Question:* What does each dot represent?

*Answer:* A sample mean budget from a sample of n=10

 *Question:* What is the shape of your sampling distribution?

*Answer:* Slightly right skewed.

*Question:* Where is your distribution centered?

*Answer:* About $53 million, which is the population mean budget. 

```{r}
mean(movies$Budget, na.rm = TRUE)
```

*Question:* The distribution should be centered at the population mean. Verify that the distribution is centered around the population mean, $\mu = 53.48$. 

*Answer:* It is very close to the population mean.

```{r}
# r-code
mean(sample.mean1000)
```


*Question:* What is the standard deviation of this distribution? (Hint: use the 95% rule.) 

*Answer:* About 15 million. 


*Question:* The standard deviation of sampling distribution has a separate name. It is called the **Standard Error**. Verify the standard deviation of this distribution using R-code.

*Answer:* It is 14.80 million quite cose to our previous informed guess.

```{r}
# r-code
sd(sample.mean1000)
```



### (e) Repeat part(d) with sample size 50 instead of 10. Generate 1000 samples.

```{r}

# Generate 1000 samples
n.size <- 50
n.rep <- 1000

Budget <- movies$Budget[!is.na(movies$Budget)] # remove NAs

sample1000 <- lapply(1:n.rep, function(i) sample(Budget, size = n.size)) 
sample.mean1000 <- lapply(sample1000, function(x) mean(x)) 
sample.mean1000 <- unlist(sample.mean1000)

mydata <- data.frame(x = sample.mean1000)

# Plot a dot plot of the sample proportion
ggplot(mydata, aes(x = sample.mean1000)) +
  geom_dotplot(dotsize=1, method = "histodot", stackratio=0.9, binwidth=1) +
  ggtitle("A single sample mean") +  xlab("Sample mean") + ylab("Count")+
  scale_x_continuous(limits = c(1,120))+
  theme(plot.title = element_text(hjust = 0.5))+
  theme(plot.title = element_text(hjust = 0.5))


```


*Question* Is this sampling distribution more or less symmetric compared to the distribution when $n=10$?

*Answer* The distribution is more symmetric with n=50 than when n=10. 

\vspace*{1in}

## Example 5: Effect of sample size

Let's investigate the effect of sample size in the sampling distribution using the same setting as in Exercise 1 with $p=0.66$. The following are three sampling distributions corresponding to different sample sizes. 


```{r, echo=FALSE}
# echo=FALSE hides this chunk of code in your R-markdown output!
pop.prop <- 0.66
data.size.25 <-  data.frame(x = rbinom(n = 200, size = 25, p = pop.prop)/25, 
                           y = "Sample Size: 25")
data.size.100 <-  data.frame(x = rbinom(n = 200, size = 100, p = pop.prop)/100, 
                           y = "Sample Size: 100")
data.size.400 <-  data.frame(x = rbinom(n = 200, size = 400, p = pop.prop)/400, 
                           y = "Sample Size: 400")
dat <-  rbind(data.size.25, data.size.100, data.size.400)

ggplot(dat, aes(x=x)) +
   geom_dotplot(dotsize=0.22,  stackratio=0.9, binwidth=0.005) + theme_bw() +
   facet_wrap(~y, ncol=1)+
   ggtitle("Effect of Sample Size") +  xlab("Proportion") + ylab("Count")+
   theme(plot.title = element_text(hjust = 0.5))
```


*Question* What happens if we increase the sample size?

*Answer* When we increase the sample size, the variability of the sampling distribution becomes smaller.


*Question* Estimate the standard error of each and verify your answer to the previous question.

*Answer* The standard errors are

```{r}
sd(data.size.25$x)
sd(data.size.100$x)
sd(data.size.400$x)
```

As the sample size increases, the variability as measured by the standard error of the sampling distribution does indeed decrease.

## Example 6: Bootstrap Sampling

```{r}
# Movies Example Again!

Budget <- movies$Budget[!is.na(movies$Budget)]

# Bootstrap samples
n.size <- 10
boot.sample1 <- sample(Budget, 10, replace = TRUE) # sampling with replacement

n.rep <- 1000
boot.sample1000 <- lapply(1:n.rep, function(i) sample(Budget, 10, replace = TRUE))
boot.samplemean1000 <- lapply(boot.sample1000, function(x) mean(x))
boot.samplemean1000 <- unlist(boot.samplemean1000)

```


```{r}
# Plot the bootstrap distribution

mydata <- data.frame(x = boot.samplemean1000)

# Plot a dot plot of the sample proportion
ggplot(mydata, aes(x = boot.samplemean1000)) +
  geom_dotplot(dotsize=0.9, stackratio=0.9, binwidth=1, method = "histodot") +
  ggtitle("Sampling distribution for sample mean") +  xlab("Sample mean") + ylab("Count")+
  scale_x_continuous(limits = c(1,120))+ 
  theme(plot.title = element_text(hjust = 0.5))

```

(a) Compare the center/spread/shape of the bootstrap distribution to the distribution computed in Ex. 4 (d). Answer all the questions in Ex. 4(d).

*Answer:* The shape/center and variability of this bootstrap distribution is very similar to that of Ex 4 (d)

```{r}
mean(mydata$x)
sd(mydata$x)
```