# Class Activity 19


## Example 1: Florida Lakes pH

The textbook dataset `FloridaLakes` contains data on 53 lakes in Florida. We want to know if the average pH of lakes in Florida is different from a neutral value of 7. 

```{r}
lakes <- read.csv("http://www.lock5stat.com/datasets1e/FloridaLakes.csv")
head(lakes)
```

#### (a) EDA 

Always plot your data and get summary stats:

```{r}
hist(lakes$pH)
mean(lakes$pH)
sd(lakes$pH)
```

- What are the sample mean and standard deviation? Use appropriate notation.
- Can we use t-inference methods with the pH variable?


<details><summary><red>Click for answer</red></summary>

*Answer:* The average pH was $\bar{x} = 6.591$ with a standard deviation of $s=1.288$. The distribution of pH is symmetric with no outliers, so we can use t-inference methods.  
</details><br>

#### (b) SE for the sample mean

What is the estimated SE for the sample mean?
<details><summary><red>Click for answer</red></summary>

*Answer:* The estimated SE for the sample mean is $SE_{\bar{x}} = 0.1770$.

```{r}
sd(lakes$pH)/sqrt(53)
```

</details><br>

#### (c) t-test statistic

Using your SE from (b) to compute the t-test statistic for testing if the population mean pH is equal, or not, to 7. Write down your hypotheses then show how the t test statistic is calculated. Interpret this value in context. 

<details><summary><red>Click for answer</red></summary>

*Answer:* The hypotheses are $H_0: \mu = 7$ vs $H_A: \mu \neq 7$. The test stat is 

$$
t = \dfrac{6.591 - 7}{1.288/\sqrt{53}} = -2.3134
$$

The observed mean of 6.591 is 2.3 SEs below the hypothesized mean of 7.

```{r}
(mean(lakes$pH) - 7)/(sd(lakes$pH)/sqrt(53)) 
```

### (d) One-sample t-test

The function `t.test(x, mu=)` can be used for a one sample test comparing the sample mean of `x` to the hypothesized value given to  `mu=`. Here we are testing whether the population mean is equal to 7 or not:

```{r}
t.test(lakes$pH, mu = 7)
```
</details><br>


- What is the t test stat given in the output? Verify that it matches your answer to (c), within reasonable rounding error.

<details><summary><red>Click for answer</red></summary>

*Answer:* The test stat  is -2.31. 

</details><br>

- What is the p-value for the test? Interpret this value.

<details><summary><red>Click for answer</red></summary>

*Answer:* The p-value is 0.025. If the mean pH of all lakes is 7, then we would see a sample mean that is at least 2.31 SEs away from 7 about 2.5% of the time in samples of 53 lakes. 

</details><br>


- What is your test conclusion?

<details><summary><red>Click for answer</red></summary>

*Answer:* There is a statistically significant difference between the observed mean pH of 6.591 and the hypothesized mean of 7 (t=-2.31, df=52, p=0.025).

</details><br>

#### (e) One-sample t confidence interval

What is the 95% confidence interval for the population mean pH? Interpret this CI. 

<details><summary><red>Click for answer</red></summary>

*Answer:* We are 95% confident that the mean pH of all lakes in Florida is between 6.24 and 6.95.

</details><br>


#### (f) `qt` and `pt`

Show how to compute the p-value for your test in (d) using the `pt` command. Then show how the confidence interval in (e) is computed with a `qt` value. 
<details><summary><red>Click for answer</red></summary>

*Answer:* For the two-sided test, the p-value is twice the proportion below the test stat $t=-2.313$ under a t-distribution with $df=53-1=52$

```{r}
2*pt(-2.313,df=52)
```

For a 95% CI, we get the 97.5th percentile from the same t-distribution

```{r}
qt(.975,52)
```

</details><br>



## Example: Nutrition Study

The dataset `NutritionStudy` contains data on daily calorie intake and other variables for 315 individuals. We want to know if the average daily calorie intake is different from the recommended 2000 calories.

```{r}
#install.packages("Lock5Data")
library(Lock5Data)
library(dplyr)
data(NutritionStudy)
glimpse(NutritionStudy)
```


### (a) EDA 

# Always plot your data and get summary stats:

```{r}
hist(NutritionStudy$Calories)
mean(NutritionStudy$Calories)
sd(NutritionStudy$Calories)
```


### (b) SE for the sample mean

# What is the estimated SE for the sample mean?


<details><summary><red>Click for answer</red></summary>

*Answer:*

```{r}
n <- length(NutritionStudy$Calories)
SE <- sd(NutritionStudy$Calories) / sqrt(n)
SE
```

</details><br>

### (c) t-test statistic

# Compute the t-test statistic for testing if the population mean calorie intake is equal, or not, to 2000.

<details><summary><red>Click for answer</red></summary>

*Answer:*

```{r}
t_stat <- (mean(NutritionStudy$Calories) - 2000) / SE
t_stat
```

</details><br>

### (d) One-sample t-test

# Perform a one-sample t-test to test whether the population mean calorie intake is equal to 2000 or not.

<details><summary><red>Click for answer</red></summary>

*Answer:*

```{r}
t.test(NutritionStudy$Calories, mu = 2000)
```

</details><br>

### (e) One-sample t confidence interval

# What is the 95% confidence interval for the population mean calorie intake?

<details><summary><red>Click for answer</red></summary>

*Answer:*

```{r}
ci <- t.test(NutritionStudy$Calories, mu = 2000)$conf.int
ci
```

</details><br>

### (f) `qt` and `pt`

# Show how to compute the p-value for the test in (d) using the pt command. Then show how the confidence interval in (e) is computed with a qt value.

<details><summary><red>Click for answer</red></summary>

*Answer:*

```{r}
p_value <- 2 * pt(-abs(t_stat), df = n - 1)
p_value
```


```{r}
t_star <- qt(0.975, df = n - 1)
t_star
```

</details><br>
