# Class Activity 20


## Example 1: Food poisoning

Suppose in an outbreak, 447 of the 998 individuals who ate beef curry were observed to have food poisoning symptoms. As researchers, suppose we want to test the hypothesis that the probability (long run proportion) of a "random individual who ate beef curry" having food poisoning is 0.1. Conduct an appropriate hypothesis test.

<details><summary><red>Click for answer</red></summary>

*Answer:*The set of hypotheses are:

$\mathrm{H}_0$ : The proportion of individuals who eat beef curry and get sick is $0.1$ : true $p=0.1 = p_o$

$\mathrm{H}_{\mathrm{a}}$ : The proportion of individuals who eat beef curry and get sick is not $0.1$ : true $p \neq 0.1$

The expected count assuming the null hypothesis is true is $n*p_0 = 998*0.1 = 99.8$. The expected count is larger than 5, so we can proceed with the chi-square test. The observed count is 447. So, the test statistics can be constructed as

\begin{align*}
\chi^2 = \frac{(O-E)^2}{E} = \frac{(447 - 99.8)^2}{99.8} = 1207.894
\end{align*}

```{r}
(447 - 99.8)^2/99.8
```

The degrees of freedom corresponding to this test is 1. So, the p-value can be calculated to be 0 as:

```{r}
1 - pchisq(1207.894, df = 1)
```

We can also do the test in R using the `prop.test` function. The results are slightly different due to continuity correction, but the results will be the same in most scenarios.


We reject the null hypothesis $(\chi^2 = 1207.894, df = 1, p-value \approx 0)$. There is a significant evidence that the proportion of individuals who eat beef curry and get sick is not $0.1$ 

</details><br>


## Example 2: Candy flavors

We have bags of candy with five flavors in each bag. We collect a random sample of ten bags. Each bag has 100 pieces of candy and five flavors. Use Chi-square goodness of fit test to test if the proportions of the five flavors in each bag are the same. The data table below shows the combined flavor counts from all 10 bags of candy. Fill in the details below:


\begin{tabular}{|c|c|c|c|c|c|} \hline
Flavor & Observed Count (O) & Expected Count (E) & $O-E$ & $(O-E)^2$ & $(O-E)^2/E$ \\  \hline
Apple & 180 & 200 & &  &  \\ \hline 
Lime & 250 & 200 & & &  \\ \hline 
Cherry & 120 & 200 & & & \\ \hline 
Orange & 225 & 200 & & &  \\ \hline 
Grape & 225 & 200 & & & \\ \hline 
\end{tabular}

<details><summary><red>Click for answer</red></summary>

*Answer:* 

$$H_0: p_1 = p_2 = p_3 = p_4 = p_5 = 1/5$$
$$H_a: \text{at least one $p_i$ not equal to 1/5}$$

```{r}
1 - pchisq(52.75, df = 5-1)
```


The observed test statistics is:

\begin{align*}
\chi^2 &= \sum_{i=1}^5 \frac{(O_i-E_i)^2}{E_i} = \frac{(180 - 200)^2}{200} + \frac{(250 - 200)^2}{200}\\ & \qquad + \frac{(120 - 200)^2}{200} + \frac{(225 - 200)^2}{200} + \frac{(225 - 200)^2}{200}\\
&=  52.75
\end{align*}


```{r}
chisq.test(x = c(180, 250, 120, 225, 225), p = rep(1/5,5))
```

We reject the null hypothesis $(\chi^2 = 52.75, df = 4, p-value \approx 0)$. We have significant evidence to claim that at least one proportion of flavors is not the same as others.

</details><br>

