# Class Activity 24


## Example 1: Data from online reviews of inkjet printers is available under the dataset `InkjetPrinters`. The data is extracted from reviews of inkjet printers at PCMag.com in August 2011. We are interested in the quantitative variable `Price`, the typical retail price of printers in dollars and `CostColor`, the cost per page in cents for printing in color.


```{r}
# load the data
inkjet <- read.csv("https://www.lock5stat.com/datasets2e/InkjetPrinters.csv")
```


```{r}
mod1 <- lm(Price~CostColor, data = inkjet)
summary(mod1)
anova(mod1)

```


```{r}
coeff <- coefficients(mod1)
eq = paste0("y = ", round(coeff[1],1), round(coeff[2],1), "*x")

ggplot(inkjet, aes(x=CostColor, y=Price))+ theme_classic()+
    geom_point()+
    geom_abline(intercept = coeff[1], slope = coeff[2], color = "red",
                linetype = "dashed", size = 1.5)+
    ggtitle(eq)


```


### (1a) What is the regression line? Please do not forget to define the variables in the regression equation.


*Answer:*

$$\widehat{Price} = 378.20 - 18.56\cdot CostColor$$

\vspace*{1in}


### (1b) What `Price` should we expect when `CostColor` is 9? 

*Answer:*

```{r}
378.20 - 18.56*9
```


\vspace*{1in}


### (1c) Find the $95\%$ confidence interval for the mean price of inkjet printers that cost 9 cents per page to print in color. You can use the `predict` function in R to calculate the confidence interval.

```{r}
# new data frame with the new x*
newx = data.frame(CostColor=9.0)
predict(mod1,newx, interval="confidence",  level = 0.95)
```


*Answer:*


\vspace*{1in}


The following is a plot of $95\%$ confidence interval for the mean price. 

```{r}
# 95% confidence interval
ggplot(inkjet, aes(x=CostColor, y=Price))+ theme_classic()+
    geom_point()+
    geom_smooth(method=lm, se=TRUE)
```

\vspace*{1in}


### (1d) Find the $95\%$ prediction interval for the price of a particular inkjet printer that cost 9 cents per page to print in color. Again, you can use the `predict` function in R to calculate the prediction interval.

```{r}
newx=data.frame(CostColor=9)
predict(mod1,newx, interval="prediction", level = 0.95)
```


*Answer:*


\vspace*{1in}


The following is a plot of $95\%$ prediction interval for price of an individual inkjet printer. 

```{r}
pred_interval <- predict(mod1, interval = "prediction")

final_df <- cbind(inkjet, pred_interval)

ggplot(final_df, aes(x=CostColor, y=Price))+ theme_classic()+
    geom_point() +
    geom_line(aes(y=lwr), color = "red", linetype = "dashed")+
    geom_line(aes(y=upr), color = "red", linetype = "dashed")+
    geom_smooth(method=lm, se=TRUE)

```



### (1e) Verify the confidence interval in part (1c) by hand. The formula for calculating the confidence interval is

$$\hat{y} \pm t^{*} s_{\epsilon} \sqrt{\frac{1}{n}+\frac{\left(x^{*}-\bar{x}\right)^{2}}{(n-1) s_{x}^{2}}} $$

```{r}
summary(inkjet$CostColor)
sd(inkjet$CostColor)
```

*Answer:*

```{r}
(211.1552 ) + c(-1,1)*2.101*sqrt(4369)*sqrt(1/20 + (9-11.67)^2/(19*sd(inkjet$CostColor)^2))
```

$$211.1552 \pm 2.101* \sqrt{4369}*\sqrt{\frac{1}{20} + \frac{(9.0 - 11.67)^2}{(20-1)2.966^2}}$$

$$(168.89, 253.42)$$

\vspace*{3in}




### (1f) Verify the prediction interval in part (1d) by hand.  The formula for calculating the prediction interval is

$$\hat{y} \pm t^{*} s_{\epsilon} \sqrt{1+\frac{1}{n}+\frac{\left(x^{*}-\bar{x}\right)^{2}}{(n-1) s_{x}^{2}}}$$

*Answer:*

```{r}
(211.1552 ) + c(-1,1)*2.101*sqrt(4369)*sqrt(1+1/20 + (9-11.67)^2/(19*sd(inkjet$CostColor)^2))
```

$$211.1552 \pm 2.101* \sqrt{4369}*\sqrt{1+\frac{1}{20} + \frac{(9.0 - 11.67)^2}{(20-1)2.966^2}}$$
$$(65.99, 356.32)$$


\vspace*{3in}


### (1g) Conduct a hypothesis test to see if we can predict the price of inkjet printers (in dollars) based on the cost per page (in cents) for printing in color (i.e., test if the linear model is effective). You can check the conditions for the test by looking at the plots provided.

*Answer:*


```{r}
inkjet$Residuals <- mod1$residuals
ggplot(inkjet, aes(x=CostColor, y=Residuals))+ theme_classic()+
geom_point()+
geom_hline(yintercept = 0, color = "red")
```


```{r}
ggplot(inkjet, aes(sample = Residuals)) + geom_qq() + geom_qq_line()+theme_classic()
```



