
# Practice Problems 4

## Problem 1: Flowers v. Mississippi

The data set `APM_DougEvensCases.csv` contains data from 1517 potential black and white jurors for 66 cases that Doug Evans was primary prosecutor for between 1992 and 2017. These jurors were available for Doug Evans to strike using his "peremptory strikes" during the jury selection phase. 


(a). Inspect data

Read in the data

```{r}
jurors <- read.csv("https://raw.githubusercontent.com/deepbas/statdatasets/main/APM_DougEvansCases.csv")
```

```{r}
# dimension of dataset
dim(jurors)
```

Look at the first **three rows** of the data set

```{r}
jurors[c(1,2,3), ]
```

To get the data from one variable, we use the command `dataset$variable`. For example, `jurors$struck_state` gives us the data values from the `struck_state` variable, which tells us if a juror was struck by the state from the jury pool. Here we can see the first 10 entries in this variable:

```{r}
jurors$struck_state[1:10]
```

<br>

(b). Table of counts and proportions

The `summary` command used with a data frame gives summaries of each variable

```{r}
summary(jurors)
```

The `table` command gives the distribution of counts for a single categorical variable. To obtain the count table for `struck_state` you need to 

```{r}
counts <- table(jurors$struck_state)
counts
```

We can add the `prop.table` command to turn these counts into proportions:

```{r}
prop.table(counts)
```

- What proportion of eligible jurors were struck by the state from the jury pool?

<details>
<summary><red>Click for answer</red></summary>
*Answer:* about 28.5% of eligible jurors were struck by the state.
</details>
<br>

(c). Bar graph for one variable

We can create a data frame `count_data` containing the counts and their corresponding categories. 

```{r}
count_data <- data.frame(counts)
count_data
```


Then, we use `ggplot2` to create a bar plot with the categories on the x-axis and the counts on the y-axis. The column names of `count_data` are automatically assigned to be `Var1` and `Freq`. We can change the column names to `category` and `count`, for example, as:

```{r}
colnames(count_data) = c("category", "count")
count_data
```


The `geom_bar(stat = "identity")` function is used to create the bars, and we set the y-axis label using `labs()`.

```{r}
# Create a bar plot using ggplot2
library(ggplot2)  # load the package
ggplot(count_data, aes(x = category, y = count)) +
  geom_bar(stat = "identity") +
  labs(y = "count")
```


You may simply use `geom_bar` if the count data is not available.


```{r}
ggplot(jurors, aes(x = struck_state)) + geom_bar() + ylab("Count")
```



(d). Two-way tables

First 10 entries of `race` and `struck_state` variable is

```{r}
jurors[(1:10),(2:3)]
```

The `table` command also gives two-way tables when two variables are included. Here is the two-way table for juror race and state struck status:

```{r}
mytable <- table(jurors$race, jurors$struck_state)
mytable
```
- How many jurors were white and were not struck by the state?

<details>
<summary><red>Click for answer</red></summary>
*answer:* 859
</details>
<br>

(e). Conditional proportions: state strike status by juror race

The `prop.table` command gives conditional proportions for a two-way table. We plug our two-way table into `prop.table` with a `margin=1` to get proportions grouped by the **row** variable:

```{r}
prop.table(mytable, margin = 1)
```

Of all eligible black jurors, about 57.9% were struck by the state.

- What proportion of eligible white jurors were struck by the state?

<details>
<summary><red>Click for answer</red></summary>
*answer:* about 12.5%
</details>

- Is there evidence of an association between juror race and state strikes?

<details>
<summary><red>Click for answer</red></summary>
*answer:* Yes, there is an association because the rate of state strikes varies greatly by juror race with
about 60% of black jurors were struck compared to only 13% of white jurors
</details><br>

(f). Stacked bar graph for two variables

We can visualize the conditional distribution from part (e) with a stacked bar graph created using the `ggplot2` graphing package. First, load this package's functions with the `library` command:

```{r}
library(ggplot2)
```

Now we can use the `geom_bar` command in this package. Here we get the conditional distribution of `struck_state` given `race`:

```{r}
ggplot(jurors, aes(x = race, fill = struck_state)) + 
  geom_bar(position = "fill") + 
  labs(title = "State strikes by juror race", y = "proportion", 
       x = "eligible juror race", fill = "struck by state?")
```

The basic syntax for this function is to let `ggplot` know your data set name (`jurors`), then specify the grouping or conditional variable on the x-axis (`race`) in the `aes` (aesthetic) argument. The `fill` variable is the response variable (`struck_state`). We add (`+`) the `geom_bar` geometry to get a bar graph with the `fill` position specified. Adding an informative label and title complete the graph.
<br>

(g). Conditional distribution of race grouped by strike status

We can "flip" our response and grouping variables easily (if we think it makes sense to do so). Here we specify the `margin=2` to get proportions grouped by the **column** variable:

```{r}
prop.table(mytable, margin = 2)
```

Notice that the proportions add to one **down** each column. Of all eligible jurors struck by the state, about 71.6% were black.

The stacked bar graph for this distribution is

```{r}
ggplot(jurors, aes(x = struck_state, fill = race)) + 
  geom_bar(position = "fill") + 
  labs(title = "Juror race by state strikes", y = "proportion", 
       fill = "eligible juror race", x = "struck by state?")
```

- What proportion of eligible jurors who were not struck by the state were black? were white?

<details>
<summary><red>Click for answer</red></summary>
*Answer:* Of all jurors not struck by the state, about 20.8% were black
</details>
<br>

---------------------------------------------------

## Problem 2: Graduate programs acceptance and sex

How are grad school program acceptance rates associated with sex? We will look at a classic data set from Berkeley grad school applications from 1973 (*Science*, 1975). The data cases are applicants to four graduate programs at Berkeley during 1973. The variable `result` tells us if the applicant was accepted to the graduate program, `sex` tells us the sex of the applicant (male or female), and  `program` tells us program type (programs 1,2,3 or 4). 

```{r}
grad <- read.csv("https://raw.githubusercontent.com/deepbas/statdatasets/main/BerkeleyGrad.csv")
```


(a). Table of counts and proportions

```{r}
prop.table(table(grad$result))
```

<!-- - What proportion of applicants were accepted? -->

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:* About 43% (1284/3014) of applicants were accepted. -->
<!-- </details> -->
<!-- <br> -->

(b). Two-way tables

The `table` command also gives two-way tables when two variables are included. Here is the two-way table for result and sex:

```{r}
table(grad$sex, grad$result)
```

<!-- - How many applicants involved females who were accepted? -->

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:* : 262 applicants involved females who were accepted. -->
<!-- </details> -->
<!-- <br> -->

(c). Conditional proportions: acceptance given sex

The `prop.table` command gives conditional proportions for a two-way table. First let's save the two-way table in an object named `mytable`:

```{r}
mytable <- table(grad$sex, grad$result)
```

Then use `prop.table` to get the distribution of result conditioned (grouped) on applicant's sex:

```{r}
prop.table(mytable, 1)
```

The value of `1` in this command tell's R that you want *row* proportions (the denominator of the proportion is each row total). 

<!-- - What proportion of female were accepted?  -->

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:* about 31% (262/(262+587)) -->
<!-- </details> -->

<!-- - What proportion of males were accepted? -->

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:* about 47% (1022/(1022+1143)) -->
<!-- </details> -->
<!-- <br> -->

(d). Bar graph for one variable

We can create a data frame `count_data1` containing the counts and their corresponding categories. 

```{r}
counts1 <- table(grad$result)
count_data1 <- data.frame(counts1)
```


Then, we use `ggplot2` to create a bar plot with the categories on the x-axis and the counts on the y-axis. The column names of `count_data` are automatically assigned to be `Var1` and `Freq`. We can change the column names to `category` and `count`, for example, as:

```{r}
colnames(count_data1) = c("category", "count")
count_data1
```


The `geom_bar(stat = "identity")` function is used to create the bars, and we set the y-axis label using `labs()`.

```{r}
# Create a bar plot using ggplot2
library(ggplot2)  # load the package
ggplot(count_data1, aes(x = category, y = count)) +
  geom_bar(stat = "identity") +
  labs(y = "count")
```

<br>

(e). Stacked bar graph for two variables

Now we can use the `geom_bar` command in this package. Here we get the conditional distribution of `result` given `sex`:

```{r}
library(ggplot2) # don't need if you already entered it for example 1
ggplot(grad, aes(x = sex, fill = result)) + 
  geom_bar(position = "fill") + 
  labs(y="Proportion", title = "result by sex", fill = "result?", x = "sex")
```

The basic syntax for this function is to let `ggplot` know your data set name (`grad`), then specify the grouping or conditional variable on the x-axis (`sex`) in the `aes` (aesthetic) argument. The `fill` variable is the response variable (`result`). We add (`+`) the `geom_bar` geometry to get a bar graph with the `fill` position specified. Adding an informative label and title complete the graph.

<br>


(f). Subsetting by program type

Finally, we will repeat the previous analysis of result and sex, but this time we will divide (or subset) the data set by program type. To do this we need to know how the values of `program` are coded:

```{r}
table(grad$program)
```

Here we use the `filter` command available from the `dplyr` package to get only the applicants to program 1:

```{r, message=FALSE}
library(dplyr)
grad.p1 <- filter(grad, program == "program1")  # gets rows where program equal program1
head(grad.p1)
dim(grad.p1)
```

- Repeat the `filter` command to get a data set for program 2 and call the new data set `grad.p2`. Verify that the number of rows in this dataset matches the number of program 2 applicants in the original data set. 


```{r}
# enter R code for (f) here
grad.p2 <- filter(grad, program == "program2") # gets rows where program equal program1
head(grad.p2)
```
<br>

(g). Result by sex for program 1.

- The distribution of result conditioned on applicant's sex for the program 1 data set is shown below. 



```{r}
ggplot(grad.p1, aes(x = sex, fill = result)) +
 geom_bar(position = "fill") +
 labs(y="Proportion", title = "result by sex for program 1",
 fill = "result?", x = "sex")
```


Get both a table of conditional proportions (or percentages) and a stacked bar graph. 

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->

<!-- ```{r} -->
<!-- prop.table(table(grad.p1$sex, grad.p1$result),1) -->
<!-- ``` -->

<!-- </details> -->
<!-- <br> -->

(h). Result by sex for program 2.

- Repeat part (g) but this time use the program 2 data set. Compare the two bar graphs for (g) and (h) and explain how they show that females have a higher acceptance rate after accounting for program type (1 or 2). 

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->

<!-- ```{r} -->
<!-- # enter R code for (h) here -->
<!-- ggplot(grad.p2, aes(x = sex, fill = result)) + -->
<!--  geom_bar(position = "fill") + -->
<!--  labs(y="Proportion", title = "result by sex for program 2", -->
<!--  fill = "result?", x = "sex") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- prop.table(table(grad.p2$sex, grad.p2$result),1) -->
<!-- ``` -->

<!-- *Answer:* For both programs 1 and 2, we see that female applicants have a slightly higher rate of acceptance -->
<!-- than male applicants. After accounting for program type, we now see that black defendants have a higher -->
<!-- rate of death penalty than white defendants. Without accounting for program type, the opposite was true -->
<!-- (see parts (c) and (e)). -->

<!-- </details> -->
<!-- <br> -->

Why? the confounding affect of program type which is associated with both result and sex:

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->

<!-- - females prefer to apply to programs 3 and 4 while males prefer programs 1 and 2 (more than 3 and -->
<!-- 4). -->
<!--   - 44% of females applied to program 3 and 40% to program 4 -->
<!--   - 38% of males applied to program 1 and 26% to program 2 -->

<!-- ```{r} -->
<!-- prop.table(table(grad$sex, grad$program), 1) -->
<!-- ``` -->
<!-- -Programs 3 and 4 were much harder to get into than programs 1 and 2 -->
<!--   - 64% of applicants to program 1 were accepted and 63% of applicants to program 2 were accepted -->
<!--   - 6% of applicants to program 4 were accepted and 34% of applicants to program 3 were accepted -->


<!-- ```{r} -->
<!-- prop.table(table(grad$program, grad$result), 1) -->
<!-- ``` -->


<!-- So since the majority of females applied to the toughest programs (as measured by acceptance rates), -->
<!-- there overall rate of acceptance was lower for females compared to males. But when we break down -->
<!-- these rates by program type, we see that females have higher acceptance rates than males (see the -->
<!-- visual in part (i)). -->

<!-- </details> -->
<!-- <br> -->

(i). A bar graph with three variables

If we simply want to graph the relationship between result and sex for each type of program, we can avoid subsetting the data by using the `facet_wrap` command in `ggplot2`. It is one simple addition to the stacked bar graph in part (e): 

```{r}
ggplot(grad, aes(x = sex, fill = result)) + 
  geom_bar(position = "fill") + 
  labs(y="Proportion", 
       title = "result by sex for each program", 
       fill = "result?", 
       x = "sex") + 
  facet_wrap(~program)
```
<br>

- Verify that this command creates side-by-side stacked bar graphs that match your graphs in parts (g) and (h) for programs 1 and 2.

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->

<!-- *Answer:* The graphs match. -->

<!-- </details> -->


