# (PART\*) Practice Problems {-}

# Practice Problem 1

## Problem 1

---------------------------------------------------

a. Run the following chunk. Comment on the output.

```{r}
example_data = data.frame(ID = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
                          Greeting = c(rep("Hello", 5), rep("Goodbye",5)),
                          Male = rep(c(TRUE, FALSE), 5),
                          Weight = runif(n=10, 50, 300))
```


<details>
<summary><red>Click for answer</red></summary>
```{r}
example_data
```

*Answer:* We see a data frame with four columns, where the first column is an `identifier` for the cases. We have information on the greeting types, whether male or not, and weight on these cases in the remaining columns.
</details>


<br>


b. What is the dimension of the dataset called 'example_data'?

<details>
<summary><red>Click for answer</red></summary>

```{r, collapse=TRUE}
dim(example_data)
nrow(example_data)
ncol(example_data)
```

*Answer:* There are 10 rows and 4 columns.
</details>

<br>

---------------------------------------------------

## Problem 2


a. Read the dataset `EducationLiteracy` from the [Lock5](https://www.lock5stat.com/datapage2e.html) second edition book.

<details>
<summary><red>Click for answer</red></summary>

```{r}
# read in the data
library(readr)
education_lock5 <- read_csv("https://www.lock5stat.com/datasets2e/EducationLiteracy.csv")
```

</details>

<br>


b. Print the header (i.e. first 6 cases by default) of the dataset in part a.

<details>
<summary><red>Click for answer</red></summary>

```{r}
head(education_lock5)
```

</details>

<br>



c. What is the dimension of the dataset in a?


<details>
<summary><red>Click for answer</red></summary>

```{r}
dim(education_lock5)
```
*Answer:* There are 188 rows and 3 columns.

</details>


<br>


d. What type of variables are `Country`, `EducationExpenditure`, and `Literacy`?

<details>
<summary><red>Click for answer</red></summary>
*Answer:* `Country` is a categorical variable. `EducationExpenditure` and `Literacy` are both quantitative variables.
</details>



<br>



e. If we would like to use education expenditure to predict the literacy rate of each countries, which variable is the explanatory variable and which one is the response?


<details>
<summary><red>Click for answer</red></summary>
*Answer:* The education expenditure is the explanatory variable, and the literacy rate is the response.
</details>


