# Practice Problems 5

## Problem 1: Sleep

This histogram shows the distribution of hours or sleep per night for a large sample of students. 

```{r, fig.height=3, fig.width=5, eval=TRUE}
library(ggplot2)
sleep <- read.csv("http://math.carleton.edu/Stats215/Textbook/SleepStudy.csv")
ggplot(sleep, aes(x=AverageSleep)) +
  geom_histogram(fill="steelblue", bins = 30) +
  labs(title = "Distribution of Sleep Hours", x = "Hours of Sleep")
```

### (a)  Estimate the average hours of sleep per night.

<details>
<summary><red>Click for answer</red></summary>
*Answer:* The mean is around 8 hours

</details>

### (b)	Use the 95% rule to estimate the standard deviation for this data.

<details>
<summary><red>Click for answer</red></summary>
*Answer:* Most of the data is between about 6 and 10, with a mean around 8 (due to the roughly symmetric
distribution). So two standard deviations is about 2 hours of sleep, making one standard deviation about
1 hours of sleep.

Let’s check the rule! Here are the actual mean and SD:

```{r}
mean(sleep$AverageSleep)
```


```{r}
sd(sleep$AverageSleep)
```

</details>

## Problem 2: Z-scores for Test Scores

The ACT test has a population mean of 21 and standard deviation of 5.  The SAT has a population mean of 1500 and a standard deviation of 325.  You earned 28 on the ACT and 2100 on the SAT. 

### (a)  Which test did you do better on? 

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:* -->

<!-- - ACT: The z-score for the score of 28 is z = (28 - 21)/5 = 1.4. -->
<!-- - SAT: The z-score for the score of 2100 is z = (2100 - 1500)/325 = 1.85. -->
<!-- - The SAT score is 1.85 standard deviations above average while the ACT score is only 1.4 standard -->
<!-- deviations above. You did better on the SAT. -->


<!-- ```{r} -->
<!-- z_ACT <- (28 - 21) / 5 -->
<!-- z_SAT <- (2100 - 1500) / 325 -->
<!-- z_ACT -->
<!-- z_SAT -->
<!-- ``` -->

<!-- </details> -->


### (b)	For each test, find the interval that is likely to contain about 95% of all test scores. 

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:* -->

<!-- - ACT: Two standard deviations is 2(5) = 10. About 95% of ACT scores are between 21 - 10 = 11 -->
<!-- and 21 + 10 = 31. This claim assumes that ACT scores follow a bell-shaped distribution. -->
<!-- - SAT: Two standard deviations is 2(325) = 650. About 95% of SAT scores are between 1500 - 650 = -->
<!-- 850 and 1500 + 650 = 2150. This claim assumes that SAT scores follow a bell-shaped distribution. -->


<!-- ```{r} -->
<!-- ACT_lower <- 21 - 2 * 5 -->
<!-- ACT_upper <- 21 + 2 * 5 -->
<!-- SAT_lower <- 1500 - 2 * 325 -->
<!-- SAT_upper <- 1500 + 2 * 325 -->
<!-- c(ACT_lower, ACT_upper) -->
<!-- c(SAT_lower, SAT_upper) -->
<!-- ``` -->

<!-- </details> -->

------------------------------------

## Problem 3: 5 number summaries

For the given vector of observations indicate whether the resulting data appear to be symmetric, skewed to the right, or skewed to the left.

 (2, 10, 15, 20, 69, 34, 23, 2, 45)

```{r}
my_vector <- c(2, 10, 15, 20, 69, 34, 23, 2, 45)
summary(my_vector)
```

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:* Skewed right. It has a longer right tail than left since max -Q3 >> Q1 - min -->

<!-- ```{r} -->
<!-- ggplot(data.frame(x=my_vector), aes(x)) + geom_boxplot() -->
<!-- ``` -->

<!-- </details> -->


## Problem 4:  Hot dog

This boxplot shows the number of hot dogs eaten by the winners of Nathan’s Famous hot dog eating contests from 2002-2011. 

```{r, fig.height=3, fig.width=5}
hotdogs <- read.csv("https://raw.githubusercontent.com/deepbas/statdatasets/main/HotDogs.csv")
ggplot(hotdogs, aes(x = "", y = HotDogs)) +
  geom_boxplot() +
  labs(title = "Number of Hot Dogs Consumed", y = "Number of Hot Dogs") 
```



### (a)  Use the boxplot to estimate the 5 number summary and IQR for this data. Verify that there are no outliers in this data.

<!-- <details> -->
<!-- <summary><red>Click for answer</red></summary> -->
<!-- *Answer:*  -->
<!-- ```{r} -->
<!-- hotdog_q1 <- quantile(hotdogs$HotDogs, 0.25) -->
<!-- hotdog_q3 <- quantile(hotdogs$HotDogs, 0.75) -->
<!-- hotdog_iqr <- IQR(hotdogs$HotDogs) -->
<!-- lower_fence <- hotdog_q1 - 1.5 * hotdog_iqr -->
<!-- upper_fence <- hotdog_q3 + 1.5 * hotdog_iqr -->

<!-- library(dplyr) -->
<!-- outliers <- filter(hotdogs, HotDogs < lower_fence | HotDogs > upper_fence) -->
<!-- outliers -->
<!-- ``` -->

<!-- </details> -->

## Problem 5: Hollywood Movies World Gross

Let's visit the `WorldGross` analysis from the Hollywood movies data set:

```{r, fig.height=3, fig.width=5}
movies <- read.csv("https://raw.githubusercontent.com/deepbas/statdatasets/main/HollywoodMovies2011.csv")
```

### (a) Draw a boxplot of `WorldGross`.

<details>
<summary><red>Click for answer</red></summary>
*Answer:* 
```{r}
ggplot(movies, aes(x = WorldGross, y = "")) +
  geom_boxplot() +
  labs(title = "World Gross of Hollywood Movies", x = "World Gross (in millions)", y ="") 
```

</details>

How many movies are identified as outliers for world gross? 

### (b) Calculating boxplot values

Use the boxplot outlier rule to find the "fence" (cutoff) between an outlier and non-outlier for `WorldGross`. Then determine the value (of `WorldGross`) that the upper "whisker" (non-outlier) extends to. 

<details>
<summary><red>Click for answer</red></summary>
*Answer:* 
```{r}
library(tidyr)
movies_no_na <- drop_na(movies)   # drop missing values
q1_world_gross <- quantile(movies_no_na$WorldGross, 0.25)
q3_world_gross <- quantile(movies_no_na$WorldGross, 0.75)
iqr_world_gross <- IQR(movies_no_na$WorldGross)
lower_fence_world_gross <- q1_world_gross - 1.5 * iqr_world_gross
upper_fence_world_gross <- q3_world_gross + 1.5 * iqr_world_gross

outliers <- filter(movies_no_na, WorldGross < lower_fence_world_gross | WorldGross > upper_fence_world_gross)
outliers
```

</details>

(c) Create a new dataset called `movies_no_outliers` that contains only the rows from `movies_no_na` where the W`orldGross` values are within the range defined by the lower and upper fences.

<details>
<summary><red>Click for answer</red></summary>
*Answer:* 
```{r}
library(dplyr)
movies_no_outliers <- filter(movies_no_na, WorldGross >= lower_fence_world_gross & WorldGross <= upper_fence_world_gross)
```

</details>


### (d) Side-by-side boxplot

We can compare boxplots of `WorldGross` across `Genre` categories:

```{r}
ggplot(movies, aes(x = Genre, y = WorldGross)) +
  geom_boxplot() +
  labs(title = "World Gross by Genre", x = "Genre", y = "World Gross (in millions)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- What does this type of graph illustrate well about the relationship between `WorldGross` and `Genre`?

<details>
<summary><red>Click for answer</red></summary>
*Answer:* Does a good job comparing median values and extremes
</details>

- What does this type of graph not illustrate well about the relationship between `WorldGross` and `Genre`?

<details>
<summary><red>Click for answer</red></summary>
*Answer:* It doesn't illustrate sample sizes well, e.g. the fantasy genre only has 2 movies in it
</details>



