
# Tutorial 3: Dive Deeper into Data Manipulation with dplyr in R


## Introduction

In this tutorial, we will introduce you to the `dplyr` package, a crucial tool in R for data manipulation. You will learn about various `dplyr` functions such as `group_by`, `summarise`, and `filter`.

---

## dplyr Essentials

### Introduction to dplyr

`dplyr` is a package for data manipulation, offering a set of useful functions to perform operations like filtering, summarizing, and grouping your data. To use `dplyr`, you'll first need to install and load the package.

```{r}
# install.packages("dplyr")   # uncomment to install
library(dplyr)
```


### Using `group_by()` Function

The `group_by()` function is used to group your data based on one or multiple columns.


```{r}
# Using inbuilt dataset 'mtcars'
data(mtcars)
head(mtcars)
```


```{r}
# Grouping by number of gears
grouped_data <- mtcars %>% group_by(gear)
```


### Summarization using `summarise()`


After grouping, `summarise()` can help to create summary statistics for each group.


```{r}
# Continuing with the 'mtcars' dataset
summary_stats <- grouped_data %>%
  summarise(
    Mean_MPG = mean(mpg),
    Max_MPG = max(mpg),
    Min_MPG = min(mpg)
  )

summary_stats
```


### Data Filtering using `filter()`

The `filter()` function filters rows based on a certain condition.

```{r}
# Filter cars with 4 gears
filtered_data <- mtcars %>% filter(gear == 4)
```

### Identifying Outliers with `filter()`

You can use `filter()` to identify outliers based on calculated fences or boundaries.


```{r}
# For demonstration, calculate lower and upper fence for mpg
Q1 <- quantile(mtcars$mpg, 0.25)
Q3 <- quantile(mtcars$mpg, 0.75)
IQR <- Q3 - Q1
lower_fence <- Q1 - 1.5 * IQR
upper_fence <- Q3 + 1.5 * IQR

outliers <- filter(mtcars, mpg < lower_fence | mpg > upper_fence)
outliers
```


### Combining All Together

By mastering these `dplyr` functions, you can significantly streamline your data manipulation tasks in R. From grouping and summarizing data to filtering out specific information, `dplyr` offers a powerful toolkit for data manipulation.

```{r}
# Combining group_by and summarise
mtcars %>% 
  group_by(gear) %>% 
  summarise(Mean_MPG = mean(mpg), N = n())

```



This tutorial covers the `group_by`, `summarise`, and `filter` functions of the `dplyr` package and demonstrates their use with the built-in dataset `mtcars`. 

