# (PART\*) Tutorials {-}

# Tutorial 1: Embark on Your Data Journey 


## Introduction

This tutorial introduces basic arithmetic operations, data types, and data manipulation functions in R.

---

## [Basic Arithmetic in R](#basic-arithmetic-in-r)
Arithmetic operations in R are straightforward, and the results are immediately displayed in the console.

```{r}
2 + 3  # Addition
5 - 2  # Subtraction
2 * 3  # Multiplication
6 / 2  # Division
```


However, to make these results accessible and manipulable for future use, you can store them in variables.

```{r}
# Performing operations and storing results in variables
addition_result <- 2 + 3  # Addition
subtraction_result <- 5 - 2  # Subtraction
multiplication_result <- 2 * 3  # Multiplication
division_result <- 6 / 2  # Division
```


Now you can use these variables later in your code

```{r}
addition_result
```


## Data Types: Vector and DataFrame

### Vector

A vector is a 1D array in R. You can create a vector using the `c()` function.

```{r}
x <- c(1, 2, 3)
x
```

#### Using Sequencing

You can also create a vector by using sequences. For example, the following code will create a vector containing numbers from 1 to 10.


```{r}
y <- 1:10
y
```


### DataFrame

A DataFrame is a 2D table in R. You can create a DataFrame using the `data.frame()` function.

```{r}
df <- data.frame(Name = c("Alice", "Bob"), Age = c(25, 30))
df
```


### Creating a Simple Dataset

You can create a simple dataset as follows:

```{r}
dataset <- data.frame(ID = 1:5, Score = c(90, 85, 88, 92, 89))
dataset
```

### Using `sample()` Function

To generate random combinations from a vector, use the `sample()` function.

```{r}
sample(x = 1:10, size = 10, replace = FALSE)
```

### Using `head()` Function

```{r}
head(dataset)
```

## Vector and Data Indexing Using `[]`

### Vector Indexing

You can index a vector using square brackets `[]`.

```{r}
x[2]  # Access the 2nd element of vector x
```

### DataFrame Indexing

You can index a DataFrame using square brackets `[]`.

```{r}
dataset[1,]  # Access the 1st row
dataset[,1]  # Access the 1st column
```

## Reading Data from a URL

To read data from a URL, you can use the `read.csv()` function, or `read_csv()` function from `readr` package.

```{r}
url <- "https://www.lock5stat.com/datasets2e/EducationLiteracy.csv"
data_from_url <- read.csv(url)
```

### Inspecting the Data

To get the first 6 rows of the data, use the following:

```{r}
head(data_from_url)
```

To get the dimensions of the data, use the following:

```{r}
dim(data_from_url)
```

## Conclusion

This tutorial covered basic arithmetic, data types, and data manipulation functions in R.