
# Tutorial 2: Dive Deeper into Data Visualization and Statistics in R


## Introduction

This tutorial aims to enhance your data manipulation and visualization skills by introducing you to ggplot2, one of the most widely-used data visualization packages in R.

---

## ggplot2 Essentials

### Introduction to ggplot

`ggplot2` is a data visualization package built upon the principles of "The Grammar of Graphics." To use ggplot, you'll first need to install and load the package.

```{r}
# install.packages("ggplot2")   # uncomment to install
library(ggplot2)
```


### Simple Bar Graph Using `geom_bar()`

The `geom_bar()` function is used to create bar graphs. A simple usage might look like:

```{r}
# Create some example data
data <- data.frame(Category = c("A", "B", "C"), Count = c(10, 60, 30))

# Create the bar graph
ggplot(data, aes(x=Category, y=Count)) +
  geom_bar(stat="identity")
```


### Adding Aesthetics with `aes()`

The `aes()` function is used to map variables to visual properties (aesthetics) of the graph.

```{r}
# Create a graph with different colors for each bar
ggplot(data, aes(x=Category, y=Count, fill=Category)) +
  geom_bar(stat="identity")
```


### Facet Plots Using `facet_wrap()`

The `facet_wrap()` function splits the data into subplots based on a factor variable.


```{r}
# Create more complex data
data2 <- data.frame(Category = c("A", "A", "B", "B", "C", "C"),
                    Count = c(10, 20, 30, 40, 50, 60),
                    Type = c("big", "small", "big", "small", "big", "small"))

# Create the facet plot
ggplot(data2, aes(x=Category, y=Count, fill=Type)) +
  geom_bar(stat="identity") +
  facet_wrap(~Type)
```


## Working with Tables and Proportions

### Using `table()` Function

The `table()` function in R generates a table of frequencies of different combinations of factor variables.

```{r}
# Frequency table
# Replicating the gender and choice data
rep_gender <- rep(c("Male", "Female", "Female", "Female", "Male"), times = 20)
rep_choice <- rep(c("No", "No", "Yes", "No", "Yes"), times = 20)

# Generating the enlarged frequency table
data <- data.frame(gender = rep_gender, choice = rep_choice)
data
```


The `rep()` function in R replicates the specified elements in the input vector for the number of times provided in the times argument.

```{r}
table_data <- table(rep_gender, rep_choice)
table_data

```


### Using `prop.table()` to Get Proportions

```{r}
# Proportion table
prop_table_data <- prop.table(table_data)
prop_table_data
```


## Combining All Together

Let's visualize this data using ggplot, geom_bar, and facet_wrap.


```{r}
ggplot(data, aes(x = gender, fill = choice)) + 
  geom_bar() + 
  labs(y="Proportion", title = "Choice by gender") + 
  scale_fill_manual(values=c("#4169E1", "#FFC72C")) +
  facet_wrap(~choice)
```




